%----------------------------------------------------------------------------
% Abstract in hungarian
%----------------------------------------------------------------------------
\chapter*{Kivonat}\addcontentsline{toc}{chapter}{Kivonat}
A természetes nyelvfeldolgozás a gépi tanulás egy olyan ága, ami természetes nyelv számítógéppel való feldolgozásához 
elengedhetetlen eszközöket nyújt, és a modern világunkban egyre inkább elterjedté válik a gép és az emberek közötti
természetes nyelven történõ kommunikációra. Ennek megvalósításához van szükségünk a szöveg tartalmi elemzésére.

A szemantikai elemzés (semantic parsing) célja, hogy nyers szöveges adathoz automatikusan
készíthessünk szemantikai reprezentációt, azaz modellezzük a szöveg jelentését. Ha a nyelvi
jelentést fogalmak irányított gráfjaival reprezentáljuk, ezeket pedig a mondat szintaktikai
szerkezetét reprezentáló fákból kell elõállítanunk, akkor a teljes feladat egyetlen komplex gráftranszformációként
definiálható.

Egy-egy ilyen elemzõ teljesítménye közvetlenül nem, csak konkrét technológiákon keresztül 
értékelhetõ, ilyen például a gépi szövegértés (Machine Comprehension) vagy a természetes nyelvi következtetés (Natural Language Inference)
, illetve tudásbázis populáció (Knowledge Base Population). A mély szemantikai elemzés elkerülhetetlen része a lexikális következtetés (lexical inference),
melynek során egy-egy szó jelentésének reprezentációjából kiindulva próbáljuk bõvíteni az egy egész frázis vagy mondat jelentését reprezentáló gráfot.

A diplomaterv témája az elõbb említett feladatok megoldása a szemantika elemzõ \texttt{4lang} \cite{Recski:2016d} rendszer segítségével, illetve ennek bõvítése
megfelelõ inferenciát megvalósító szabályokkal és a gráfok felett értelmezett metrikákkal. A dolgozat magába fogalja a 4lang rendszer lehetõségeinek
REST API-t megvalósító mikroszolgáltatásokba való csomagolását is. 

A gépi szövegértés feladaton a kidolgozott rendszerünk elõzetes eredményei kis javulást mutatnak a 2018-ban bemutatott state-of-the art rendszerhez \cite{Wang:2018} képest.
\vfill

%----------------------------------------------------------------------------
% Abstract in english
%----------------------------------------------------------------------------
\chapter*{Abstract}\addcontentsline{toc}{chapter}{Abstract}

Natural language processing (NLP) is a branch of machine learning that provides us tools for analyzing raw text automatically. 
In our modern world human-machine communication has become a widely required task. In order to meet it's requirements, it is inevitable to analyze
the text's meaning.

The main task of semantic parsing is to automatically build semantic representation from the input, so we can model the meaning
of raw texts. If we model meaning as directed graphs of concept and we can build them from syntax trees that represents the structure of 
sentences, then we can define the whole process as one complex graph transformation.

The performance of such analyzers cannot be measured directly, only through concrete tasks, such as Machine comprehension, 
Natural language inference, or Knowledge base population. Lexical inference has a very strong connection to deep semantic parsing, 
where we want to augment the graph that represents the meaning of a concept or a sentence by taking each word's graph that models its meaning. 

The main focus of this thesis is to give a strong baseline for the task mentioned above, and to enchahnt existing systems
with the help of the semantic parser \texttt{4lang} \cite{Recski:2016d}. We believe
the significance of these experiments lies in its
demonstration that inference based graph transformations are a powerful method for solving any semantic parsing related task.
The thesis includes the improving of the 4lang system with further inference based rules and various metrics related to semantic graphs. Furthermore it will demonstrate
the wrapping the 4lang software in Restful Wep Api microservices for easier usage.   

The biggest result of this paper is achieved on the Machine Comprehension task, where we integrated our method into a state-of-the art system. Our preliminary results suggest .5 percentage point improvement.
\vfill

